---
title: "Lecture 10 - 10-3"
output: html_document
---

```{r}
library(tidyverse)
#import weather data
weather_df = 
  rnoaa::meteo_pull_monitors(c("USW00094728", "USC00519397", "USS0023B17S"),
                      var = c("PRCP", "TMIN", "TMAX"), 
                      date_min = "2016-01-01",
                      date_max = "2016-12-31") %>%
  mutate(
    name = recode(id, USW00094728 = "CentralPark_NY", 
                      USC00519397 = "Waikiki_HA",
                      USS0023B17S = "Waterhole_WA"),
    tmin = tmin / 10,
    tmax = tmax / 10,
    month = lubridate::floor_date(date, unit = "month")) %>%        #to create a 'month' variable 
  select(name, id, date, month, everything())
```

#Group_by
- 'group_by()' to group vars by categorical vars before performing subsequent functions
- 'summarise()' to create one-number summaries within each group 
- 'mutate()' to define vars within groups
```{r}
weather_df %>%
  group_by(name, month)       #can use 'ungroup_by()' to remove groupings
weather_df
```

#Counting things 

```{r}
weather_df %>%
  count(month)              #count # of elements within a group, e.g. 93 in Janurary 2016 (not limited to any particular park name)

weather_df %>%
  group_by(name, month) %>%
  summarize(n = n())        #counts the # of elements within a group within another group, e.g. 31 observations in january 2016 in central park specifically

weather_df %>%
  group_by(month) %>%
  summarize(n_obs = n(),
            n_days = n_distinct(date))  #to find the # of distinct 'date' values within each month 
```

#General Summaries

```{r}
#group by one var 
weather_df %>%
  group_by(month) %>%
  summarize(n = n(),
            mean_tmax = mean(tmax),
            mean_prec = mean(prcp, na.rm = TRUE),  #can't take mean of a var if it has even a single missing datapoint, so use 'na.rm = TRUE' to remove all the missing vars 
            median_tmax = median(tmax),
            sd_tmax = sd(tmax))

#other functions: mean(), median(), var(), sd(), mad(), IQR() = interquartile range, min(), and max()

#Group by more than one var
weather_df %>%
  group_by(name, month) %>%
  summarize(number = n(), 
            mean_tmax = mean(tmax, na.rm = TRUE),
            median_prcp = median(prcp, na.rm = TRUE),
            sd_tmin = sd(tmin, na.rm = TRUE))
```
#plot two tibbles at once
```{r}
weather_df %>%
  group_by(name, month) %>%
  summarize(mean_tmax = mean(tmax)) %>%
  ggplot(aes(x = month, y = mean_tmax, color = name)) + 
    geom_point() + geom_line() + 
    theme(legend.position = "bottom")
```

#tidy is not always best 
```{r}
weather_df %>%
  group_by(name, month) %>%
  summarize(mean_tmax = mean(tmax, na.rm = TRUE)) %>% 
  spread(key = name, value = mean_tmax) %>%     #untidy dataset in a long format by changing it to a short format so its easier to work with. Remember, key = the new column name 
  knitr::kable(digits = 3)
```

#Grouped mutate
```{r}
weather_df %>%
  group_by(name, month) %>%
  mutate(mean_tmax = mean(tmax, na.rm = TRUE))
               #mutate within the dataset that's already been grouped by name. Add a column that takes the mean of tmax within each park name 
 

weather_df %>%
  group_by(name) %>%
  mutate(mean_tmax = mean(tmax, na.rm = TRUE),      #
         centered_tmax = tmax - mean(tmax)) %>%     #mutate within the dataset that's already been grouped by name. Add a column that takes the mean of tmax within each park name. Also add a column called 'centered_tmax' that relates extreme temp maxes to the mean.  
  ggplot(aes(x = date, y = centered_tmax, color = name)) + 
    geom_point() 
```

#Window functions
if ask it to compute mean of 30 numbers, it'll spit out 30 numbers (one value repeated 30 times)
```{r}
weather_df %>%
  group_by(name, month) %>%
  mutate(tmax_rank = min_rank(tmax))         #what is the ordering of values within a specific group. Useful if want to know the coldest or warmest day in a certain month

weather_df %>%
  group_by(name, month) %>%
  mutate(tmax_rank = min_rank(tmax)) %>%
  filter(min_rank(tmax) < 2)          #can see coldest day (rank = 1) within each month so ask for < 2

weather_df %>%
  group_by(name, month) %>%
  mutate(tmax_rank = min_rank(tmax)) %>%
  filter(min_rank(desc(tmax)) < 4)      # keep the three days with the highest tmax

weather_df %>%
  group_by(name) %>%
  mutate(temp_change = tmax - lag(tmax))    #use lags to compare an obs to the previous value. E.g. to find daily change in tmax within each station within a year

weather_df %>%
  group_by(name) %>%
  mutate(one_day_change = tmax - lag(tmax)) %>%
  summarize(sd_one_day_change = sd(one_day_change, na.rm = TRUE)) #find day to day variability at each of the three stations. now, can use to compare among the stations
```

#Limitations
summarise() can only be used with functions that return single-number summary, so can't use for linear regressions and more complex functions

#Learning Asessment
In the PULSE data, the primary outcome is BDI score; it’s observed over follow-up visits, and we might ask if the typical BDI score values are roughly similar at each. Try to write a code chunk that imports, cleans, and summarizes the PULSE data to examine the mean and median at each visit. 
```{r}

group_by(BDI, visit)
```

#Learning assessment
In the FAS data, there are several outcomes of interest; for now, focus on post-natal day on which a pup is able to pivot. Two predictors of interest are the dose level and the day of treatment. 
```{r}
```

